CUBE: An Information-optimized parallel Cosmological $N$-body Algorithm

### 2018-06-03 ms_caf/ms_caf.tex
nu vtf

\documentclass[10pt,twocolumn,preprint]{emulateapj}
\usepackage[english]{babel}
\usepackage{amsmath,color,aas_macros}

@@

\affil{
$^1$Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240, China\\
$^2$Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON M5H 3H8 Canada\\
$^3$Department of Physics and Astronomy, Shanghai Jiao Tong University, Shanghai, 200240, China\\
$^4$Dunlap Institute for Astronomy and Astrophysics, University of Toronto, Toronto, ON M5S 3H4, Canada\\
$^5$Canadian Institute for Advanced Research, CIFAR Program in Gravitation and Cosmology, Toronto, ON M5G 1Z8, Canada\\
$^6$Perimeter Institute for Theoretical Physics, Waterloo, ON N2L 2Y5, Canada
}

@@

\begin{abstract}
Cosmological large scale structure $N$-body simulations are computation-light, memory-heavy problems in supercomputing. The considerable amount of memory is usually dominated by an inefficient way of storing more than sufficient phase space information of particles. We present a new parallel, information-optimized, particle-mesh based $N$-body code CUBE, in which information- and memory-efficiency is increased by nearly an order of magnitude. This is accomplished by storing particle's relative phase space coordinates instead of global values, and in the format of fixed point as light as 1-byte. The remaining information is given by complimentary density and velocity fields (negligible in memory space) and proper ordering of particles (no extra memory). Our numerical experiments show that this information-optimized $N$-body algorithm provides accurate results within the error of the particle-mesh algorithm. This significant lowering of memory-to-computation ratio breaks the bottleneck of scaling up and speeding up large cosmological $N$-body simulations on multi-core and heterogenous computing systems.

@@

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations, so many algorithms were designed to alleviate it. Various fast multipole methods \citep{1985JCoPh..60..187R} improve the complexity to $o(N\log N)$ even $o(N)$, among which the most popular one is ``tree'', like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}. Other methods include adaptive grid algorithms like {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh (PM) codes \citep{1995ApJS..100..269P}. The standard PM algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} is first codes to use a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.

@@

\section{Method}\label{s.method}
The most memory consuming part of a $N$-body simulation is usually the phase space coordinates of $N$-body particles -- 24 byte per particle (here after we use ``bpp'' referring to ``byte per particle''), which contains 4 single precision floating numbers, must be used to store each particles' 3-dimensional position and velocity vectors. {\tt CUBEP3M}, an example of a memory-efficient parallel $N$-body code, can use as low as 40 bpp in sacrificing computing speed \citep{2013MNRAS.436..540H}. This includes the phase coordinates (24 bpp) for particles in physical domain and buffered region, a linked list (4 bpp), and a global coarse mesh and local fine mesh. Sometimes 4-byte real numbers are not necessarily adequate in representing the {\it global} coordinates in simulations. If the box size is many orders of magnitude larger than interactive distance between particles, especially in the field of re-simulation of dense subregions, double precision (8-byte) coordinates are needed to avoid truncation errors. Another solution is to record {\it relative} coordinates for both position and velocity. {\tt CUBE} replaces the coordinates and linked list 24+4=28 bpp memory usage with an integer based storage, reduces the basic memory usage from 28 bpp down to 6 bpp, described as following \ref{ss.position} and \ref{ss.velocity}. The algorithm is described in \ref{ss.overview}.

@@

In a simulation with volume $L^3$ and $N_c^3$ coarse cells, particle positions are stored with a resolution of $L/(256N_c)$. The force calculation (e.g. softening length) should be configured much finer than this resolution, discussed in later sections. On the other hand, particle position can also be stored as 2-byte (16 bits) integers to increase the resolution. In this case, each coarse cell is divided into $2^{16}=65536$ bins and the position resolution is $L/(65536N_c)$, precise enough compared to using4-byte global coordinates, see later results. We denote this case ``x2'' and denote using 1-byte integers for positions ``x1''.

We collectively write the general position conversion formulae
\begin{equation}\label{eq.chi}
	\chi_d=\left[2^{8n_\chi}(x_d-\left[x_d\right])\right]-2^{8n_\chi-1},
\end{equation}
\begin{equation}\label{eq.x}
	x_d=(n_c-1)+2^{-8n_\chi}\left(\chi_d+2^{8n_\chi-1}+1/2\right),
\end{equation}
where $[\ ]$ is the operator to take the integer part. $n_\chi\in\{1,2\}$ is the number of bytes used for each integer, $x_d$ and $\chi_d$ are floating and integer versions of the coordinate. The velocity counterpart of them are $n_\nu=1,2$, $v_d$ and $\nu_d$. 
%$n_c=1,2,...$ is the coarse grid index, given by the ordering of particles and a integer based particle density field (see \ref{sss.spatial_decomposition}).
The position resolution for $n_\chi$-byte integer, ``x$n_\chi$'', is $2^{-8n_\chi}L/N_c$.

@@

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ relative to this field: 
\begin{equation}
	v_d=v_c+\Delta v.
\end{equation}
$v_c$ is always recorded and kept updated, and should not occupy considerable memory. We then divide velocity space $\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate which $\Delta v$ bin the particle is located.

@@

The essential parameter of the velocity distribution is its variance. On non-linear scale, the velocity distribution function is non-Gaussian. However, to the first order approximation, we simply assume it as Gaussian and characterized by the variance
\begin{equation}\label{eq.vdisp}
	\sigma_v(a,k) = \frac{1}{3}  (a H f D)^2 \int_0^kd^3q\frac{P_{L}(q)}{q^2},
\end{equation}
where $a(z)$ is the scale factor, $H(z)$ the Hubble parameter, $D$ is the linear growth factor, $f=d \ln D/d\ln a$, and $P_L(k)$ is the linear power spectrum of density contrast at redshift zero. $\sigma_v(a,k)$ is a function of cosmic evolution $a$ and a smoothing scale $k$, or $r$ (see Figure \ref{fig.vdisp}). $\Delta v$ is the velocity dispersion relative to the coarse grid, so we approximate its variance as
\begin{equation}\label{eq.vdelta}
	\sigma^2_{\Delta}(a)=\sigma^2_v(a,r_c)-\sigma^2_v(a,r_p), 
\end{equation}
where $r_c$ is the scale of coarse grid, and $r_p$ is the scale of average particle separation. In each dimension of 3D velocity field, we use $\sigma^2_{\Delta}(a)/3$ according to the equipartition therorem. On different scales, we measure the statistics of $v_d$, $v_c$ and $\Delta v$ and find good agreement with the above model.

@@

Initially, particles are generated by initial condition generator, at a higher redshift. The coarse grid density field $v_c$ is also generated at this step by averaging all particles in the coarse cell. A global $\sigma_\Delta$ is calculated by equation (\ref{eq.vdelta}), where linear approximation is hold. Then velocities are stored by equation (\ref{eq.nu}). During the simulation, $v_c$ is updated every time step, and a nonlinear $\sigma_\Delta$ is measured directly from the simulation, and can be simply used in the next time step, after scaled by the ratio of growth factors between two adjacent time steps. More details see section \ref{sss.algorithm}.

@@

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} decomposes the global simulation volume into $M_g^3$ cubic sub-volumes with $N_c$ coarse grids or $N_f=RN_c$ fine grids per side. The fine-mesh is usually $R=4$ times finer than the coarse mesh. Each of these sub-volumes is assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into $M_t^3$ cubic {\it tiles} (defined as $V_p$, or {\it physical} region) with $N_c/(M_g M_t)$ coarse grids per side. Each $V_p$ is surrounded by a {\it buffer} region $V_b$ which is $N_b$ coarse cells thick. We define extended tile region $V_e\equiv V_p+V_b$. $V_e$ is designed for two reasons: (1) if the short fine-mesh force ${\bs F}_f$ has a cut-off $N_b$, and is computed by $\rho_c$ whose domain $D(\rho_f)=V_e$, then ${\bs F}_f$ in $V_p$ is guaranteed to be correct, i.e.,
\begin{eqnarray}\label{eq.cond1}
\nonumber	{\bs F}_f(r>N_b)&=&0 \wedge D(\rho_c)=V_e\\
	&\Rightarrow& \{{\bs F}_f(\rho_c) |{\bs F}_f\in V_p \} {\rm \ is\ correct},
\end{eqnarray}
and (2) to collect all particles being able to travel to $V_p$, i.e.,
\begin{equation}\label{eq.cond2}
	{\bs v}\,{\rm d}t \leq N_b \Rightarrow \{ {\bs x} | {\bs x}+{\bs v}\,{\rm d}t\in V_p\} \in V_e.
\end{equation}
Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with $M_g=2$ and $M_t=2$.

According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,v_c,\chi_d,\nu_d\}$ by using Fortran language:
\begin{eqnarray*}
	&&{\tt integer(1)}\ \ \, \rho_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt real(4)}\ \ \ \ \ \ \ \, v_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt integer(}n_\chi{\tt )}\, \chi_d(3,P_{\rm max})[M_g,M_g,*]\\
	&&{\tt integer(}n_\nu{\tt )}\ \nu_d(3,P_{\rm max})[M_g,M_g,*]
\end{eqnarray*}
where $N_e=N_t+2N_b$ covers the buffer region on both sides, $M_t$ is the tile dimemsions, and $M_g$ is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image $P_{\rm local}$, and $P_{\rm max}>P_{\rm local}$ is a value (discussed in Section \ref{ss.memory}) large enough to store particles in $V_e$. $\chi_d$ and $\nu_d$
%(equivalent to $\chi_d$ and $\nu_d$ in \ref{ss.position} and \ref{ss.velocity} respectively)
are required to be sorted according to the same memory layout of $\rho_c$, such that $n_c$ and $x_d$ can be obtained by equation (\ref{eq.x}). 
$\{\rho_c,v_c,\chi_d,\nu_d\}$ provides a complete information of positions and velocities of particles, and we call it a checkpoint.

@@

The second step to generate particles and displace them by Zel'dovich approximation (ZA) \citep{1970A&A.....5...84Z}, where the displacement field is obtained by differentiating $\Phi$, ${\bs \Psi}({\bs q})=\nabla\Phi({\bs q})$ in real space. This step is done on $V_e$ of each tile. Similar to Equation (\ref{eq.cond2}), we have
\begin{equation}\label{eq.cond3}
	\max(\Psi_d) \leq N_b \Rightarrow \{ {\bs q} | {\bs q}+{\bs \Psi}({\bs q})\in V_p\} \in V_e.
\end{equation}
We iterate twice over particles' ${\bs q}$ in $V_e$. The first iteration calculates particles' ${\bs x}$ and ${\bs v}$ by ZA and obtains $\rho_c$ and $v_c$ on the coarse grid. The second iteration, ${\bs x}$ and ${\bs v}$ is calculated again and are converted to $\chi_d$ and $\nu_d$ by Equations (\ref{eq.chi},\ref{eq.nu}) and place them in a certain order according to $\rho_c$. Lastly, we delete particles in $V_b$, and resort the ones in $V_p$ and write $\{\rho_c,v_c,\chi_d,\nu_d,I_P({\rm optional})\}$ of this tile to disk. The above is similar to {\tt update\_xp} of Section \ref{sss.algorithm}. If PIDs are needed, they are also generated here.
After working on all tiles, we sum up $P_{\rm local}$ and $P_{\rm global}$. We summarize the above steps into a pseudocode in Figure \ref{fig.ic}. During this step, the only major memory usage is $\Phi$ on the fine mesh. If the number of particles per fine grid $P_f=1$, the memory consumption of this in-place FFT is 4 bpp.

@@

{\tt timestep} --
Runge-Kutta 2 method is used in the time integration. i.e., for $n$ time steps, we update positions ($D$=drift) and velocities ($K$=kick) at interlaced half time-steps by operator splitting; the operation $(DKKD)^n$ is 2nd order accurate. The actual simulation applies varied time steps by {\tt timestep}, where a time increment ${\rm d}t$ is constrained by particles' maximum velocities, accelerations, cosmic expansion and any other desired conditions.

@@

{\tt update\_xp} --
According to ${\rm d}t$, {\tt update\_xp} updates particle positions (drift $D$) in a ``{\it gather}'' algorithm
%(in contrast, {\tt CUBEP3M} uses ``scatter'' algorithm)
tile by tile. For each particle, $\chi_d$ and $\nu_d$ are converted to $x_d$ and $v_d$ by Equations(\ref{eq.x},\ref{eq.v}). Due to Condition (\ref{eq.cond2}), we are able to collect all particles to travel to $V_p$.

@@

{\tt update\_vp} -- 
the PM or particle-particle particle-mesh (P$^3$M) algorithm is applied in this subroutine to update particles's velocities (kick $K$).
We first call {\tt buffer\_density} and {\tt buffer\_xp} in order that particle positions are in buffered state. Then according to the particle distributions $\rho_f$ on $V_e$ we calculate the fine-mesh force ${\bs F}_f$ and update the particle velocities in the $V_p$. Condition (\ref{eq.cond1}) guarantees the correctness of this step.  An optional particle-particle (PP) force ${\bs F}_{\rm pp}$ can be called to increase the force resolution.

@@

By far, $\nu_d$ in physical regions are correctly updated. We simply call {\tt buffer\_v} again to bring $\nu_d$ into buffered state, such that the {\tt update\_x} in the next iteration will be done correctly.

{\tt checkpoint} -- 
If a desired redshift is reached, we execute the last drift step in the $(DKKD)^n$ operation by {\tt update\_xp}, and call {\tt checkpoint} to save the disjoint state of $\{\rho_c,v_c,\chi_d,\nu_d\}$ on disk. Related operations, like run-time halo finder, projections are also done at this point.
%If final desired redshift is reached, {\tt final\_step} let us exit the main loop. These corresponding logical variables are controlled in {\tt timestep}.

@@

\subsubsection{Compare with traditional algorithms}
To illustrate the improvement of memory usage, we refer to the TianNu simulation \citep{2017NatAs...1E.143Y} run on the Tianhe-2 supercomputer, which used a traditional $N$-body code {\tt CUBEP3M}. TianNu's particles number is limited by memory per computing node -- for each computing node, an average of $576^3$ neutrino particles and $288^3$ CDM particles are used, and consumes $\mathcal{M}=40\,{\rm GB}$\footnote{Additional memory is used for OpenMP parallelization and for particle IDs to differentiate different particle spices.}, or about $\mathcal{M}_P=186$. A memory efficient configuration of {\tt CUBEP3M} by using large physical scales and at costs of speed, still uses about $\mathcal{M}_P=40$.

@@

\begin{figure*}
\centering
  \includegraphics[width=1.0\linewidth]{f9}
 \caption{Dimensionless power spectra $\Delta^2(k)$ and decorrelations $\xi(k)$ (lower axes) induced by the integer-based algorithms in simulations S2048S and S2048L. The four solid lines show the results from \{x2v2,x1v2,x2v1,x1v1\}. In lower panels, the dashed curves show the decorrelations given by the PM algorithm, measured by cross correlating different resolutions of {\tt CUBE}-x2v2. The vertical orange dashed lines show the scale $k=0.2\, k_{\rm Nyquist}$.}
\label{fig.ccc}
\end{figure*}
\subsection{Power spectrum}

@@

For x2v2, almost all particles are accurate up to 1/100 of a fine grid, and the worst particle is $\sim 1/10$ of a fine grid away from its counterpart in {\tt CUBEP3M}. The difference is caused by truncation errors and is negligible in physical and cosmological applications. The accuracy of x1v2 is between x2v2 and x1v1, and x2v1 gives only minor improvement from x1v1. We also run a simulation with same number of particles but with $L=600\ {\rm Mpc}/h$ ($m_p=1.2\times 10^{12}\Msun$), and find that the accuracy of x1v2 is in turn between x2v1 and x1v1. We interpret that, in this latter case, particles have lower mass resolution, so they move slower and need higher position resolution but need lower velocity resolution, then x2v1 outperforms x1v2.

>>>>

\documentclass[10pt,twocolumn,preprint]{emulateapj}
\usepackage[english]{babel}
\usepackage{amsmath,color,aas_macros,ulem}

@@

\affil{
$^1$Tsung-Dao Lee Institute, Shanghai Jiao Tong University, Shanghai, 200240, China\\
$^2$Canadian Institute for Theoretical Astrophysics, University of Toronto, Toronto, ON M5H 3H8 Canada\\
$^3$Department of Astronomy, Shanghai Jiao Tong University, Shanghai, 200240, China\\
$^4$Dunlap Institute for Astronomy and Astrophysics, University of Toronto, Toronto, ON M5S 3H4, Canada\\
$^5$Canadian Institute for Advanced Research, CIFAR Program in Gravitation and Cosmology, Toronto, ON M5G 1Z8, Canada\\
$^6$Perimeter Institute for Theoretical Physics, Waterloo, ON N2L 2Y5, Canada
}

@@

\begin{abstract}
Cosmological large scale structure $N$-body simulations are computation-light, memory-heavy problems in supercomputing. The considerable amount of memory is usually dominated by an inefficient way of storing more than sufficient phase space information of particles. We present a new parallel, information-optimized, particle-mesh based $N$-body code CUBE, in which information- and memory-efficiency is increased by nearly an order of magnitude. This is accomplished by storing particle's relative phase space coordinates instead of global values, and in the format of fixed point as light as 1-byte. The remaining information is given by complimentary density and velocity fields (negligible in memory space) and proper ordering of particles (no extra memory). Our numerical experiments show that this information-optimized $N$-body algorithm provides accurate results within the error of the particle-mesh algorithm. This significant lowering of memory-to-computation ratio breaks the bottleneck of scaling up and speeding up large cosmological $N$-body simulations on multi-core and heterogeneous computing systems.

@@

When $N$ is large, the brute force pairwise particle-particle (PP) force brings unaffordable $o(N^2)$ computations, so many algorithms were designed to alleviate it. Various fast multipole methods \tcb{\citep{1985JCoPh..60..187R,2014ComAC...1....1D,2016ascl.soft09016P}} improve the complexity to $o(N\log N)$ even $o(N)$, among which the most popular one is ``tree'', like {\tt GADGET} \citep{2001NewA....6...79S,2005MNRAS.364.1105S} and its simulation ``Millennium'' \citep{2005Natur.435..629S,2012MNRAS.426.2046A}, {\tt TPM} \citep{1995ApJS...98..355X} and {\tt GOTPM} \citep{2004NewA....9..111D}. Other methods include adaptive grid algorithms like {\tt HYDRA} \citep{1995ApJ...452..797C} and {\tt RAMSES} \citep{2010ascl.soft11007T}, as well as mesh-refined codes \citep{1991ApJ...368L..23C} and moving adaptive particle-mesh (PM) codes \citep{1995ApJS..100..269P}. The standard PM algorithm \citep{1988csup.book.....H} is most memory and computational efficient if we focus on large cosmological scales. The load balancing problem is minimized as the matter distribution is rather homogeneous, and the speed benefits from the fast Fourier transform (FFT) libraries, such as {\tt FFTW3} \citep{Frigo05thedesign}. {\tt PMFAST} \citep{2005NewA...10..393M} introduces a 2-level PM algorithm, aiming to put PM codes toward speed, memory compactness and scalability. After subsequent developments on {\tt PMFAST}, {\tt CUBEP3M} \citep{2013MNRAS.436..540H} uses cubic spatial decomposition, and adds PP force and many other features.

@@

\section{Method}\label{s.method}
The most memory consuming part of a $N$-body simulation is usually the phase space coordinates of $N$-body particles -- 24 byte per particle (here after we use ``bpp'' referring to ``byte per particle''), which contains 6 single precision floating numbers, must be used to store each particles' 3-dimensional position and velocity vectors. {\tt CUBEP3M}, an example of a memory-efficient parallel $N$-body code, can use as low as 40 bpp in sacrificing computing speed \citep{2013MNRAS.436..540H}. This includes the phase coordinates (24 bpp) for particles in physical domain and buffered region, a linked list (4 bpp), and a global coarse mesh and local fine mesh. Sometimes 4-byte real numbers are not necessarily adequate in representing the {\it global} coordinates in simulations. If the box size is many orders of magnitude larger than interactive distance between particles, especially in the field of re-simulation of dense subregions, double precision (8-byte) coordinates are needed to avoid \tcb{round-off} errors. Another solution is to record {\it relative} coordinates for both position and velocity. {\tt CUBE} replaces the coordinates and linked list 24+4=28 bpp memory usage with an integer based storage, reduces the basic memory usage from 28 bpp down to 6 bpp, described as following \ref{ss.position} and \ref{ss.velocity}. The algorithm is described in \ref{ss.overview}.

@@

In a simulation with volume $L^3$ and $N_c^3$ coarse cells, particle positions are stored with a resolution of $L/(256N_c)$. The force calculation (e.g. softening length) should be configured much finer than this resolution, discussed in later sections. On the other hand, particle position can also be stored as 2-byte (16 bits) integers to increase the resolution. In this case, each coarse cell is divided into $2^{16}=65536$ bins and the position resolution is $L/(65536N_c)$, precise enough compared to using 4-byte global coordinates, see later results. We denote this case ``x2'' and denote using 1-byte integers for positions ``x1''.

We collectively write the general position conversion formulae
\begin{equation}\label{eq.chi}
	\chi_d=\left[2^{8n_\chi}(x_d-\left[x_d\right])\right]-2^{8n_\chi-1},
\end{equation}
\begin{equation}\label{eq.x}
	x_d=(n_c-1)+2^{-8n_\chi}\left(\chi_d+2^{8n_\chi-1}+1/2\right),
\end{equation}
where $[\ ]$ is the operator to take the integer part. $n_\chi\in\{1,2\}$ is the number of bytes used for each integer, $x_d$ and $\chi_d$ are floating and integer versions of the coordinate. The velocity counterpart of them are $n_\nu=1,2$, $v_d$ and $\nu_d$.
The position resolution for $n_\chi$-byte integer, ``x$n_\chi$'', is $2^{-8n_\chi}L/N_c$.

@@

\subsection{Particle velocity storage}\label{ss.velocity}
Similarly, actual velocity in $d$th dimension $v_d$ is decomposed into an averaged velocity field on the same coarse grid $v_c$ and a residual $\Delta v$ relative to this field:
\begin{equation}\label{eq.deltav}
	v_d=v_c+\Delta v.
\end{equation}
$v_c$ is always recorded and kept updated, and should not occupy considerable memory. We then divide velocity space $\Delta v$ into uneven bins, and use a $n_\nu$-byte integer to indicate which $\Delta v$ bin the particle is located.

@@

The essential parameter of the velocity distribution is its variance. On non-linear scale, the velocity distribution function is non-Gaussian. However, to the first order approximation, we simply assume it as Gaussian and characterized by the variance
\begin{equation}\label{eq.vdisp}
	\tcb{\sigma^2_v(a,r) = (a H f D)^2 \int_0^\infty d^3k\frac{P(k)}{k^2}W^2(k,r)},
\end{equation}
where $a(z)$ is the scale factor, $H(z)$ is the Hubble parameter, $D$ is the linear growth factor, $f=d \ln D/d\ln a$, and $P(k)$ is the linear power spectrum of density contrast at redshift zero. \tcb{$W(k,R)$ is the Fourier transform of real space top-hat window function with smoothing scale $r$. In Figure \ref{fig.vdisp} we plot $\sigma_v(a,r)$ as a function of $a$ for a few smoothing scale $r$. $\Delta v$ in equation (\ref{eq.deltav})} is the velocity dispersion relative to the coarse grid, so we approximate its variance as
\begin{equation}\label{eq.vdelta}
	\sigma^2_{\Delta}(a)=\sigma^2_v(a,r_c)-\sigma^2_v(a,r_p),
\end{equation}
where $r_c$ is the scale of coarse grid, and $r_p$ is the scale of average particle separation. In each dimension of 3D velocity field, we use $\sigma^2_{\Delta}(a)/3$ according to the equipartition therorem. On different scales, we measure the statistics of $v_d$, $v_c$ and $\Delta v$ and find good agreement with the above model.

@@

Initially, particles are generated by initial condition generator, at a higher redshift. The coarse grid \tcb{velocity} field $v_c$ is also generated at this step by averaging all particles in the coarse cell. A global $\sigma_\Delta$ is calculated by equation (\ref{eq.vdelta}), where linear approximation is hold. Then velocities are stored by equation (\ref{eq.nu}). During the simulation, $v_c$ is updated every time step, and a nonlinear $\sigma_\Delta$ is measured directly from the simulation, and can be simply used in the next time step, after scaled by the ratio of growth factors between two adjacent time steps. More details see section \ref{sss.algorithm}.

@@

\subsubsection{Spacial decomposition}\label{sss.spatial_decomposition}
{\tt CUBE} decomposes the global simulation volume into $M_g^3$ cubic sub-volumes with $N_c$ coarse grids or $N_f=RN_c$ fine grids per side. The fine-mesh is usually $R=4$ times finer than the coarse mesh. Each of these sub-volumes is assigned to a coarray {\it image}. Inside of an image, the sub-volume is further decomposed into $M_t^3$ cubic {\it tiles} (defined as $V_p$, or {\it physical} region) with $N_c/(M_g M_t)$ coarse grids per side. Each $V_p$ is surrounded by a {\it buffer} region $V_b$ which is $N_b$ coarse cells thick. We define \tcb{the} extended tile region \tcb{as} \tcb{$V_e\equiv V_p\cup V_b$}. $V_e$ is designed for two reasons:

(1) if the short fine-mesh force ${\bs F}_f$ has a cut-off, \tcb{$N_b$ ${\bs F}_f(r>N_b)=0$}, and is computed \tcr{\sout{ by $\rho_c$ whose domain $D(\rho_f)=V_e$}} \tcb{on $V_e$}, then ${\bs F}_f$ in $V_p$ is guaranteed to be correct.

\tcb{(2) $V_e$ is able to collect all particles that are able to travel to $V_p$.} \tcr{\sout{i.e., ... }[deleted previous two equations (8,9)]}
%\begin{eqnarray}\label{eq.cond1}
%\nonumber	{\bs F}_f(r>N_b)&=&0 \wedge D(\rho_c)=V_e\\
%	&\Rightarrow& \{{\bs F}_f(\rho_c) |{\bs F}_f\in V_p \} {\rm \ is\ correct},
%\end{eqnarray}

%and (2) to collect all particles being able to travel to $V_p$, i.e.,
%\begin{equation}\label{eq.cond2}
%	{\bs v}\,{\rm d}t \leq N_b \Rightarrow \{ {\bs x} | {\bs x}+{\bs v}\,{\rm d}t\in V_p\} \in V_e.
%\end{equation}
Figure \ref{fig.tile} shows the spacial decomposition in a 2-dimensional analogy, with $M_g=2$ and $M_t=2$.

According to this spatial decomposition, and as discussed in the last two subsections, we declare $\{\rho_c,v_c,\chi_d,\nu_d\}$ by using Fortran language:
\begin{eqnarray*}
	&&{\tt integer(1)}\ \ \, \rho_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt real(4)}\ \ \ \ \ \ \ \, v_c(N_e,N_e,N_e,M_t,M_t,M_t)[M_g,M_g,*]\\
	&&{\tt integer(}n_\chi{\tt )}\, \chi_d(3,P_{\rm max})[M_g,M_g,*]\\
	&&{\tt integer(}n_\nu{\tt )}\ \nu_d(3,P_{\rm max})[M_g,M_g,*]
\end{eqnarray*}
where $N_e=N_t+2N_b$ covers the buffer region on both sides, $M_t$ is the tile dimemsions, and $M_g$ is the image {\it codimensions}\footnote{Coarray Fortran concept. Codimensions can do communications between images.}. We denote actual number of particles in a given image $P_{\rm local}$, and $P_{\rm max}>P_{\rm local}$ is a value (discussed in Section \ref{ss.memory}) large enough to store particles in $V_e$. $\chi_d$ and $\nu_d$
are required to be sorted according to the same memory layout of $\rho_c$, such that $n_c$ and $x_d$ can be obtained by equation (\ref{eq.x}).
$\{\rho_c,v_c,\chi_d,\nu_d\}$ provides a complete information of positions and velocities of particles, and we call it a checkpoint.

@@

The second step to generate particles and displace them by Zel'dovich approximation (ZA) \citep{1970A&A.....5...84Z}, where the displacement field is obtained by differentiating $\Phi$, ${\bs \Psi}({\bs q})=\nabla\Phi({\bs q})$ in real space. This step is done on $V_e$ of each tile. \tcr{\sout{Similar to condition (9), we have ...} [deleted previous equation (10).]}
%\begin{equation}\label{eq.cond3}
%	\max(\Psi_d) \leq N_b \Rightarrow \{ {\bs q} | {\bs q}+{\bs \Psi}({\bs q})\in V_p\} \in V_e.
%\end{equation}

We iterate twice over particles' ${\bs q}$ in $V_e$. The first iteration calculates particles' ${\bs x}$ and ${\bs v}$ by ZA and obtains $\rho_c$ and $v_c$ on the coarse grid. The second iteration, ${\bs x}$ and ${\bs v}$ is calculated again and are converted to $\chi_d$ and $\nu_d$ by Equations (\ref{eq.chi},\ref{eq.nu}) and place them in a certain order according to $\rho_c$. Lastly, we delete particles in $V_b$, and resort the ones in $V_p$ and write $\{\rho_c,v_c,\chi_d,\nu_d,I_P({\rm optional})\}$ of this tile to disk. The above is similar to {\tt update\_xp} of Section \ref{sss.algorithm}. If PIDs are needed, they are also generated here.
After working on all tiles, we sum up $P_{\rm local}$ and $P_{\rm global}$. We summarize the above steps into a pseudocode in Figure \ref{fig.ic}. During this step, the only major memory usage is $\Phi$ on the fine mesh. If the number of particles per fine grid $P_f=1$, the memory consumption of this in-place FFT is 4 bpp.

@@

{\tt timestep} --
\tcb{The second order Runge-Kutta method} is used in the time integration. i.e., for $n$ time steps, we update positions ($D$=drift) and velocities ($K$=kick) at interlaced half time-steps by operator splitting; the operation $(DKKD)^n$ is 2nd order accurate. The actual simulation applies varied time steps by {\tt timestep}, where a time increment ${\rm d}t$ is constrained by particles' maximum velocities, accelerations, cosmic expansion and any other desired conditions.

@@

{\tt update\_xp} --
According to ${\rm d}t$, {\tt update\_xp} updates particle positions (drift $D$) in a ``{\it gather}'' algorithm tile by tile. For each particle, $\chi_d$ and $\nu_d$ are converted to $x_d$ and $v_d$ by Equations(\ref{eq.x},\ref{eq.v}). \tcr{\sout{Due to Condition (9), we are able to collect all particles to travel to $V_p$.}}

@@

{\tt update\_vp} --
the PM or particle-particle particle-mesh (P$^3$M) algorithm is applied in this subroutine to update particles's velocities (kick $K$).
We first call {\tt buffer\_density} and {\tt buffer\_xp} in order that particle positions are in buffered state. Then according to the particle distributions $\rho_f$ on $V_e$ we calculate the fine-mesh force ${\bs F}_f$ and update the particle velocities in the $V_p$. \tcr{\sout{Condition (8) guarantees the correctness of this step.}}  An optional particle-particle (PP) force ${\bs F}_{\rm pp}$ can be called to increase the force resolution.

@@

\tcb{After the updating of $\nu_d$ in $V_p$, we} simply call {\tt buffer\_v} again to bring $\nu_d$ into buffered state, such that the {\tt update\_x} in the next iteration will be done correctly.

{\tt checkpoint} --
If a desired redshift is reached, we execute the last drift step in the $(DKKD)^n$ operation by {\tt update\_xp}, and call {\tt checkpoint} to save the disjoint state of $\{\rho_c,v_c,\chi_d,\nu_d\}$ on disk. Related operations, like run-time halo finder, projections are also done at this point.

@@

\subsubsection{Compare with traditional algorithms}
To illustrate the improvement of memory usage, we refer to the TianNu simulation \citep{2017NatAs...1E.143Y} run on the Tianhe-2 supercomputer, which used a traditional $N$-body code {\tt CUBEP3M}. TianNu's particle number \tcb{(shown in Table \ref{t.sim})} is limited by memory per computing node -- for each computing node, an average of $576^3$ neutrino particles and $288^3$ CDM particles are used, and consumes $\mathcal{M}=40\,{\rm GB}$\footnote{Additional memory is used for OpenMP parallelization and for particle IDs to differentiate different particle spices.}, or about $\mathcal{M}_P=186$. A memory efficient configuration of {\tt CUBEP3M} by using large physical scales and at costs of speed, still uses about $\mathcal{M}_P=40$.

@@

\begin{figure*}
\centering
  \includegraphics[width=1.0\linewidth]{f9}
 \caption{Dimensionless power spectra $\Delta^2(k)$ and \tcb{the cross-correlations $\xi(k)$ (lower axes) with respect to x2v2} in simulations S2048S and S2048L. The four solid lines show the results from \{x2v2,x1v2,x2v1,x1v1\}. In lower panels, the dashed curves show the decorrelations given by the PM algorithm, measured by cross correlating different resolutions of {\tt CUBE}-x2v2. The vertical orange dashed lines show the scale $k=0.2\, k_{\rm Nyquist}$.}
\label{fig.ccc}
\end{figure*}
\subsection{Power spectrum}

@@

For x2v2, almost all particles are accurate up to 1/100 of a fine grid, and the worst particle is $\sim 1/10$ of a fine grid away from its counterpart in {\tt CUBEP3M}. The difference is caused by \tcb{round-off} errors and is negligible in physical and cosmological applications. The accuracy of x1v2 is between x2v2 and x1v1, and x2v1 gives only minor improvement from x1v1. We also run a simulation with same number of particles but with $L=600\ {\rm Mpc}/h$ ($m_p=1.2\times 10^{12}\Msun$), and find that the accuracy of x1v2 is in turn between x2v1 and x1v1. We interpret that, in this latter case, particles have lower mass resolution, so they move slower and need higher position resolution but need lower velocity resolution, then x2v1 outperforms x1v2.
